An Efficient Minibatch Acceptance Test for MCMC

### 2016-06-21 nips_2016.tex
update by JFC

\title{A Simple Minibatch Acceptance Test for MCMC}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on the full data. To do this, we
describe a novel acceptance test which is applied to minibatches of
data. Previous work on minibatch Metropolis-Hastings use adaptive
batch sizes, which may use far more than one minibatch of data and therefore
yield few samples.  Here we describe a simpler test on fixed-sized
minibatches. Instead of adaptive batch sizes, our test requires a
precondition on the variance of the log proposal probability
ratio. Given a measurement of this variance, a minibatch size can be
chosen to support the test in one step, which yields a number of
samples proportional to the batch/minibatch ratio. Alternatively, the
variance condition can be used to adjust the temperature of the
sampling distribution to a ``natural'' value to provide regular
samples at a given minibatch size. The resulting proposal/test is only
slightly more complex than a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and therefore have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations: 
GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed, etc.

In this paper, we develop an alternative minibatch acceptance test for MCMC
methods which uses only one fixed-size minibatch per test. Instead of
dynamically adjusting the batch size, we add a prerequisite for the test
based on the variance of the log acceptance probability. If this
variance is small enough, the test requires only a single step. This
variance changes slowly on typical problems, and we can estimate it
before performing the test. The minibatch variance
decreases as the minibatch size grows, so the variance measurement can be
used to set the minimum minibatch size.

There is another advantage when using our test for Bayesian posterior inference. As the dataset size
grows, the posterior distribution sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become stuck in one of them. The
samples are not exploring the posterior parameter space, and are unlikely to concentrate at a strong
local optimum. They therefore fail to accurately represent a sample of the posterior because of
inadequate search (i.e., mixing). We argue that it is much more natural to sample the distribution
at a higher temperature determined by the minibatch size rather than the dataset size. Starting at a
high temperature (small minibatch size) which mixes easily, one can reduce the temperature
(increasing minibatch size) until one reaches the desired target distribution temperature.  Although
this approach also involves changing minibatch size, the changes increase minibatch size
monotonically, and only a few changes are performed in the course of a sampling session.  Annealing
can also be used for MAP and ML estimation because changing temperature preserves posterior modes.

To be precise, the contributions of this paper are as follows:

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch acceptance test which satisfies detailed balance.
    \item We compare performance of our new test and the adaptive sampling method.
    \item We experiment using the approach for posterior estimation.
\end{enumerate}

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. This $P_a$ satisfies detailed balance, so if
one samples long enough, one will arrive at a stationary distribution matching the posterior, though
a burn-in period and/or only using samples at regular intervals is often done in practice.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on a probability ratio. If the test cannot make a decision
over a certain confidence threshold, then the minibatch size is increased and the test repeats. This
process continues until a decision. The downside of this algorithm is the need to keep incrementing
the minibatch size; in the worst case, all $N$ data points may be needed to decide on $\theta'$. We prove that the lower bound of average number of data instance consumed for one iteration is $O(\sqrt{N})$, where $N$ is the number of all data points, see Lemma~\ref{lem:efficienty_cutting}. A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties
based on concentration inequalities, but comes at the cost of more computational time.

\begin{lemma}\label{lem:efficienty_cutting}
If the sample mean $\bar{l}$ has uniform distribution from $[-L,L]$, the lower bound of expected data consumed of adaptive sampling method for one iteration is $\frac{\eta s_{l}Q}{2L}\sqrt{N}$, where $\bar{l} = \frac{1}{n}\sum_{i=1}^{n} \left (\log p(x_i; \theta') - \log p(x_i| \theta_t)\right )$, $\eta = \phi^{-1}_{n-1}(1-\varepsilon)$, $Q=\frac{p(\theta_t)q(\theta'|\theta_t)}{p(\theta')q(\theta_t|\theta')}$, $s_l$ is the population standard deviation of $l_i$, $\phi_{n-1}()$ is the CDF of t distribution with degree of freedom $n-1$, and $n$ is the minibatch size.
\end{lemma}

The work of~\cite{conf/uai/MaclaurinA14} presents another minibatch MCMC method which gets the
\emph{exact} posterior, so long as there exists a (cheap) lower bound on the likelihood for each
data point. Their approach is only a starting point, however, and it can be difficult to derive
these lower bounds in general cases. The extra auxiliary variables they introduce also hurt mixing.

There is a separate line of MCMC work on simulating the physics of a probability distribution. By
viewing random variables as particles in a system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate the best case scenario for proposals: high
quality \emph{and} distant.  HMC methods require a full gradient computation of the posterior, which
means they also face the problem of computing the likelihood at each point. Stochastic versions of
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use a minibatch of data each iteration but require
controlling the minibatch noise to avoid divergence. Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a similar approach which does not use the
momentum terms in HMC.  Minibatch SGHMC and Langevin Dynamics are orthogonal to our work and are
other ways of attempting to solve the same problem of MCMC methods on big data. We can combine our
methods with these results, as we demonstrate in Section~\ref{ssec:nets}.

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^N$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
MAP/ML estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\;
    Compute the $\Delta'$ from one minibatch data\;
    Compute the ${\rm std}(\Delta')$ from another minibatch data\;
    \While {${\rm std}(d) \ge 1.2$} {
    	Update the value of $\Delta'$ and ${\rm std}(\Delta')$ from one new minibatch\;
    }

        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;

        \eIf{$X_c + \Delta' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
        }

>>>>

\title{An Efficient Minibatch Acceptance Test for MCMC}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data reductions
and in the worst case require the entire dataset. 
Here we present a simple test which requires constant size minibatches
if an auxiliary condition is met. We describe several cases under which
this condition can be satisfied. Our new test is naturally tolerant
of ``noisy'' estimates of the transition probability ratio. To achieve
this, we replace the classical Metropolis-Hastings test with a smooth
acceptance function which is a sigmoid (a logistic function). 
The resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and therefore have high sample efficiency. They also
map very well onto hardware such as GPUs. In contrast, MCMC methods
typically require calculations over the full dataset to produce a new
sample.  Recent results
from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
(bounded error) acceptance tests based on the classical M-H test. The
tests depend on minibatch statistics, and on the value of an
additional random variable. The amount of data consumed for each test
varies significantly from one minibatch to the next, and
non-repeatably on the random variable. By contrast,
\cite{conf/uai/MaclaurinA14} performs exact tests but requires a lower
bound on parameter distribution across its domain. The amount of data
reduction depends on the accuracy of this bound, and such bounds are
only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. Our test requires a bound on the
(relative) variance of the log acceptance probability ratio. This variance
changes slowly on typical problems, and we can estimate it before
performing the test.

If the variance test fails, there are three
ways to mitigate the situation:
\begin{itemize}

\item Increase the temperature of the target distribution. Log likelihoods
  scale as $1/T$, and so the variance of the likelihood ratio will
  vary as $1/T^2$.

\item Increase the minibatch size. Relative variance scales as $1/k$ vs the
  minibatch size $k$.

\item For continuous probability distributions, reduce the proposal step
  size. Since the condition is on variance in the {\em difference} in
  log probabilities of prior and proposed likelihoods, smaller proposal steps
  will lead to smaller changes in probability and smaller changes in the
  variance of those changes.
\end{itemize}

Note that changes to the distribution temperature may only need to be
temporary.  The distribution can be annealed from high-to-low
temperatures to maintain the variance condition. High temperature is
natural with small minibatches and large step sizes, since the
precision of the acceptance test is limited by the amount of data
available. As the distribution reaches a stationary state and is cooled,
particle motion can be slowed with larger minibatches or smaller proposal
steps. The contributions of this paper are as follows:

\begin{itemize}
\item We develop a new minibatch acceptance test which satisfies detailed balance.
\item we developed effectively measurable bounds for both absolute and relative
  test error. 
\item We compare performance of our new test and prior approaches. 
\item We experiment using the approach for posterior estimation.
\end{itemize}

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the samples.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. This $P_a$ satisfies detailed balance, so if
one samples long enough, one will arrive at a stationary distribution matching the posterior, though
a burn-in period and/or only using samples at regular intervals is often done in practice.

Unfortunately, computing $f$ requires the use of all $N$ training data
points. Moreover, it is difficult to design tests using substantially
fewer than $N$ points that also satisfy detailed balance. To reconcile
these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and tests the hypothesis that the sample $\theta'$ should be
accepted or rejected based on a probability ratio involving the
additional random variable $u\in [0,1]$. If the test cannot make a
decision with sufficient confidence, then the minibatch size is
increased and the test repeats. This process continues until a
decision. The bounds depend on either a central limit theorem
\cite{cutting_mh_2014} or a concentration bound
\cite{icml2014c1_bardenet14} requiring direct bounds on the log
likelihood ratio.  The problem with both methods is the expense of
resolving small differences.  In the worst case, all $N$ data points
may be needed, and as we show next this worst case can occur
about $\sqrt{N}$ times during the performance of $N$ tests.

We next prove that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014,icml2014c1_bardenet14} is
$\Omega(\sqrt{N})$, where $N$ is the number of data points, see
Lemma~\ref{lem:efficienty_cutting}.

\begin{lemma}\label{lem:efficienty_cutting}
If the sample mean $\bar{l}$ has uniform distribution on $[-L,L]$,
the lower bound of expected data consumed of adaptive sampling method
for one iteration is $\frac{\eta s_{l}Q}{2L}\sqrt{N}$, where $\bar{l}
= \frac{1}{n}\sum_{i=1}^{n} \left (\log p(x_i; \theta') - \log p(x_i|
\theta_t)\right )$, $\eta = \phi^{-1}_{n-1}(1-\varepsilon)$,
$Q=\frac{p(\theta_t)q(\theta'|\theta_t)}{p(\theta')q(\theta_t|\theta')}$,
$s_l$ is the population standard deviation of $l_i$, $\phi_{n-1}()$ is
the CDF of t distribution with degree of freedom $n-1$, and $n$ is the
minibatch size.
\end{lemma}

\begin{proof} see appendix
  \end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more agressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
MAP/ML estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution\;
    Draw a minibatch of $m$ samples, compute $\Delta'(\theta',\theta_t)$ and ${\rm std}(\Delta')$ for this sample\;
    \While {${\rm std}(\Delta') \ge 1.2$} {
    	Draw $m$ more samples to augment the current minibatch, and update $\Delta'$ and ${\rm std}(\Delta')$ \;
    }
    Let $\sigma = \sqrt{1.2 - {\rm std}(\Delta')^2}$, draw a normal variable $X_n \in Norm(0,\sigma)$ and
    $X_c$ from the correction distribution\;

        \eIf{$X_c + X_n + \Delta' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old sample, $\theta_{t+1} = \theta_t$\;
        }
